The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
assorted touch-ups in Section 5, still various TODOs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2552 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 093b0a157596e2b471cab114ffb16029144d5208

## HCL palette
pal <- qualitative_hcl(5, "Dark 3")
names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE_scores", "CTree_max")

@@

@@

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the first scenario (stump with 1 split) the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of data sets for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach).
For the second scenario (tree with 2 splits) the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

To illustrate the effects of the building blocks in the unifying framework presented in 
Section~\ref{sec:unify} different combinations of them are investigated for the first
simulation scenario (stump) in Sections ~\ref{sec:scores} and \ref{sec:3way}. 
Then the testing strategies are evaluated on the more
complex tree structure of the second scenario in Section~\ref{sec:tree}. 
The corresponding results are shown in the following plots.

\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
\label{sec:scores}
To assess the impact of using only residuals or the full score matrix four testing
strategies are compared: CTree, GUIDE, MOB (all three in their original form),
and additionally the GUIDE strategy using the full score matrix (denoted by GUIDE\_scores).

\begin{figure}
\setkeys{Gin}{width=\linewidth}
<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE_scores", "GUIDE"))

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Proportion of selecting"~Z[1]),
                      par.settings = par.settings),

               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
                                                     expression("vary"~beta[0]), 
                                                     expression("vary"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
                                   bg = "gray90"))

@
\caption{\label{fig:scores_stump}Proportion of cases where the true split variable $Z_1$ is
selected over increasing effect size $\delta$ (100 data sets of 250 observations per step),
once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
Each column represents one scenario for the coefficients: 
varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
both coefficients varying (right). The performances of four testing strategies are compared: 
CTree, MOB, GUIDE, GUIDE\_scores.}
\end{figure}

Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
As explained before, the tests are applied as significance tests with significance 
level $\alpha=0.05$. Therefore, the proportion of cases in which the true split variable $Z_1$
is detected with a $p$-value smaller than $\alpha$ is compared.
(To show that in general the drawn conclusions do not change if no significance level is included 
the same setting is evaluated again without any significance level and the results are presented
in Appendix~\ref{app:significance}.)

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

@@

Altogether these results clearly illustrate that overall the decision to include all available
scores when testing for dependencies in order to select split variables is advantageous over
considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
the mean but also in all other parameters can be detected in that way. This is particularly obvious
when comparing GUIDE and GUIDE\_scores. Moreover, from the fact that these two version are both
performing worse than MOB and CTree it can be deduced that categorizing split variables and
binarizing scores has a negative effect on the ability to detect the right split variable. 
To elaborate these effects more in detail the results for all possible combinations of the 
building blocks are discussed in a full factorial 3-way analysis.

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building Block         & Levels              & Effect \\
\hline
Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
                       & \textit{scores}     & Use full score matrix\\
Categorization         & \textit{cat}        & Categorize split variables\\
                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
Binarization           & \textit{bin}        & Binarize scores\\
                       & \textit{lin}        & Use residuals/score values without binarization\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
slope parameter as it only considers residuals in its testing strategy.
For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
methods and both GUIDE versions fail to find the correct tree structure.
Based on these results it can be concluded once again that including all available scores in the
testing strategy leads to a better performance while binarizing scores and categorizing
split variables both do not cause any improvement but in many situations even have a negative
effect on the power of a testing strategy.

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
                                        "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE_scores", "GUIDE"))

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

xyplot(ari ~ delta | xi, groups = ~ test, 
       data = subdata, 
       type = "l",
       lty = c(1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(subdata$test)),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       #index.cond = list(c(1,2,3,4)),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings)

@
\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 data sets of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
MOB, GUIDE, GUIDE\_scores.}
\end{figure}

@@

prex$test <- factor(prex$test,
                    levels = c("ctree","mfluc",
                               "guide_sum_12", "guide_sum_1_cor",
                               "ctree_p","mfluc_p",
                               "guide_sum_12_p", "guide_sum_1_cor_p"),
                    labels = c("CTree","MOB",
                               "GUIDE_scores", "GUIDE",
                               "CTree_p","MOB_p",
                               "GUIDE_scores_p", "GUIDE_p"))

for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
  if(prex$test[i] == "GUIDE_scores") prex$test[i] <- "GUIDE_scores_p"
  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
}

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 data sets of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). 
The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
The dashed lines represent the results from applying pre-pruning and the continuous lines 
the results from applying post-pruning.}
\end{figure}

@@

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
and \cite{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.

@@

\newpage
\section*{Appendix}
\begin{appendix}

\section{Combinations of building blocks applied}\label{app:combinations}
For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``\_res'' refers to using only residuals (instead of all scores),
``\_scores'' to using all scores (instead of only residuals),
``\_cat'' to applying a categorization of split variables,
``\_bin'' to applying a binarization of scores/residuals.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & bin          & cat \\
MOB\_bin\_res     & residuals          & bin          & max \\
CTree\_bin\_res   & residuals          & bin          & lin \\
MOB\_cat\_res    & residuals          & lin          & cat \\
MOB\_res          & residuals          & lin          & max \\
CTree\_res        & residuals          & lin          & lin \\
GUIDE\_scores     & scores             & bin          & cat \\
MOB\_bin          & scores             & bin          & max \\
CTree\_bin        & scores             & bin          & lin \\
MOB\_cat          & scores             & lin          & cat \\
MOB               & scores             & lin          & max \\
CTree             & scores             & lin          & lin \\
\hline
\end{tabular}
\caption{Testing strategies applied (representing the 12 combinations
of building blocks).
\label{tab:combinations_appendix}}
\end{table}

\section{Significance level}\label{app:significance}
In Section~\ref{sec:scores} the investigated testing strategies are compared based
on the number of cases in which the true split variable $Z_1$ is detected with a $p$-value
smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
is due to the aim of investigating how well the tests perform as significance test.
However, ignoring this significance level yields the same conclusions 
which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
and \ref{fig:scores_stump}.
In Figure~\ref{fig:scores_stump_wosign} the proportion of cases 
in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
the exact same simulation setting is applied with a significance level.

\begin{figure}[t]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE_scores", "GUIDE"))

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

               strip =strip.custom(factor.levels = c(expression("vary"~beta[0]~"and"~beta[1]), 
                                                     expression("vary"~beta[0]), 
                                                     expression("vary"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
                                   bg = "gray90"))

>>>>

## HCL palette
pal <- qualitative_hcl(5, "Dark 3")
names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE+scores", "CTree+max")

@@

\vspace*{0.4cm}

@@

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the first scenario (stump with 1 split) the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of replications for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach).
For the second scenario (tree with 2 splits) the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

% To illustrate the effects of the building blocks in the unifying framework presented in 
% Section~\ref{sec:unify} different combinations of them are investigated for the first
% simulation scenario (stump) in Sections~\ref{sec:scores} and~\ref{sec:3way}. 
% Then the testing strategies are evaluated on the more
% complex tree structure of the second scenario in Section~\ref{sec:tree}. 
% The corresponding results are shown in the following plots.

\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
\label{sec:scores}

To assess the impact of using only residuals or the full score matrix four testing
strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
the full scores instead (denoted by GUIDE+scores).

\begin{figure}
\setkeys{Gin}{width=\linewidth}
<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Proportion of selecting"~Z[1]),
                      par.settings = par.settings),

               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
                                                     expression("varying"~beta[0]), 
                                                     expression("varying"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
                                   bg = "gray90"))

@
\caption{\label{fig:scores_stump}Proportion of replications where the true split variable $Z_1$ is
selected over increasing effect size $\delta$ (100 replications of 250 observations per step),
once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
Each column represents one scenario for the coefficients: 
varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
both coefficients varying (right). The performances of four testing strategies are compared: 
CTree, MOB, GUIDE, GUIDE+scores.}
\end{figure}

Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
As explained before, the tests are applied as significance tests with significance 
level $\alpha=0.05$. Therefore, the proportion of replications in which the true split variable $Z_1$
is detected with a $p$-value smaller than $\alpha$ is compared.
(To show that in general the drawn conclusions do not change if no significance level is included 
the same setting is evaluated again without any significance level and the results are presented
in Appendix~\ref{app:significance}.)

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE+scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

@@

Altogether these results clearly illustrate that overall the decision to include all available
scores when testing for dependencies in order to select split variables is advantageous over
considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
the mean but also in all other parameters can be detected in that way. This is particularly obvious
when comparing GUIDE and GUIDE+scores. Moreover, from the fact that these two version are both
performing worse than MOB and CTree it can be deduced that categorizing split variables and
binarizing scores has a negative effect on the ability to detect the right split variable. 
To elaborate these effects more in detail the results for all possible combinations of the 
building blocks are discussed in a full factorial 3-way analysis.

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block         & Levels              & Effect \\
\hline
Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
                       & \textit{scores}     & Use full score matrix\\
Categorization         & \textit{cat}        & Categorize split variables\\
                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
Binarization           & \textit{bin}        & Binarize scores\\
                       & \textit{lin}        & Use residuals/score values without binarization\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE+scores performs
slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
slope parameter as it only considers residuals in its testing strategy.
For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
methods and both GUIDE versions fail to find the correct tree structure.
Based on these results it can be concluded once again that including all available scores in the
testing strategy leads to a better performance while binarizing scores and categorizing
split variables both do not cause any improvement but in many situations even have a negative
effect on the power of a testing strategy.

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<tree_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_tree, test %in% c("ctree","mfluc",
                                        "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE+scores", "GUIDE"))

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))

xyplot(ari ~ delta | xi, groups = ~ test, 
       data = subdata, 
       type = "l",
       lty = c(1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(subdata$test)),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       #index.cond = list(c(1,2,3,4)),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings)

@
\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
MOB, GUIDE, GUIDE+scores.}
\end{figure}

@@

prex$test <- factor(prex$test,
                    levels = c("ctree","mfluc",
                               "guide_sum_12", "guide_sum_1_cor",
                               "ctree_p","mfluc_p",
                               "guide_sum_12_p", "guide_sum_1_cor_p"),
                    labels = c("CTree","MOB",
                               "GUIDE+scores", "GUIDE",
                               "CTree_p","MOB_p",
                               "GUIDE+scores_p", "GUIDE_p"))

for(i in c((length(pr$test)+1) : (2*length(pr$test)))){
  if(prex$test[i] == "CTree") prex$test[i] <- "CTree_p"
  if(prex$test[i] == "MOB") prex$test[i] <- "MOB_p"
  if(prex$test[i] == "GUIDE+scores") prex$test[i] <- "GUIDE+scores_p"
  if(prex$test[i] == "GUIDE") prex$test[i] <- "GUIDE_p"
}

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). 
The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE+scores. 
The dashed lines represent the results from applying pre-pruning and the solid lines 
the results from applying post-pruning.}
\end{figure}

@@

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
and \citealp{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.

@@

\bibliography{ref.bib}

\newpage

\begin{appendix}

\section{Combinations of building blocks applied}\label{app:combinations}

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``+res'' refers to using only residuals (instead of all scores),
``+scores'' to using all scores (instead of only residuals),
``+cat'' to applying a categorization of split variables,
``+bin'' to applying a binarization of scores/residuals.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l l }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & bin          & cat \\
MOB+bin+res     & residuals          & bin          & max \\
CTree+bin+res   & residuals          & bin          & lin \\
MOB+cat+res    & residuals          & lin          & cat \\
MOB+res          & residuals          & lin          & max \\
CTree+res        & residuals          & lin          & lin \\
GUIDE+scores     & scores             & bin          & cat \\
MOB+bin          & scores             & bin          & max \\
CTree+bin        & scores             & bin          & lin \\
MOB+cat          & scores             & lin          & cat \\
MOB               & scores             & lin          & max \\
CTree             & scores             & lin          & lin \\
\hline
\end{tabular}
\caption{Testing strategies applied (representing the 12 combinations
of building blocks).
\label{tab:combinations_appendix}}
\end{table}

\section{Significance level}\label{app:significance}

\begin{figure}[t]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree", "CTree+max", "MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
                                                     expression("varying"~beta[0]), 
                                                     expression("varying"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
                                   bg = "gray90"))
